---
title: "Quantium Retail Analytics - Chip Category Analysis"
author: "Olamilekan Olaiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(lubridate)
library(janitor)
library(scales)

# Load raw data
transaction_data <- read_excel("QVI_transaction_data.xlsx")
customer_data <- read_csv("QVI_purchase_behaviour.csv")

# Clean transaction data
transaction_data <- transaction_data %>%
  mutate(DATE = as.Date(DATE, origin = "1899-12-30")) %>%
  distinct()

dip_products <- c(
  "Old El Paso Salsa   Dip Tomato Mild 300g",
  "Doritos Salsa       Medium 300g",
  "Old El Paso Salsa   Dip Chnky Tom Ht300g",
  "Woolworths Mild     Salsa 300g",
  "Old El Paso Salsa   Dip Tomato Med 300g",
  "Woolworths Medium   Salsa 300g",
  "Doritos Salsa Mild  300g"
)

transaction_data <- transaction_data %>%
  filter(!PROD_NAME %in% dip_products) %>%
  filter(LYLTY_CARD_NBR != 226000) %>%
  mutate(
    BRAND = word(PROD_NAME, 1),
    BRAND = case_when(
      BRAND == "Dorito"   ~ "Doritos",
      BRAND == "Smith"    ~ "Smiths",
      BRAND %in% c("RRD", "Red") ~ "Red Rock Deli",
      BRAND == "Infzns"   ~ "Infuzions",
      BRAND == "Snbts"    ~ "Sunbites",
      BRAND %in% c("NCC", "Natural") ~ "Natural Chip Co",
      BRAND %in% c("GrnWves", "Grain") ~ "Grain Waves",
      BRAND == "WW"       ~ "Woolworths",
      TRUE ~ BRAND
    ),
    PACK_SIZE = parse_number(str_extract(PROD_NAME, "[0-9]+[gG]"))
  )

# Merge and add remaining features
merged_data <- transaction_data %>%
  left_join(customer_data, by = "LYLTY_CARD_NBR") %>%
  mutate(
    PRICE_PER_UNIT = TOT_SALES / PROD_QTY,
    SEGMENT = paste(LIFESTAGE, PREMIUM_CUSTOMER, sep = " - ")
  )
```
```

## 1. Business Problem
Julia, the Category Manager for chips, needs to understand customer 
purchasing behaviour to inform the chip category strategy for the 
next half-year.

## 2. Data Cleaning Summary
- Removed 1 duplicate row
- Removed 15,166 non-chip salsa dip products
- Removed 1 outlier commercial customer (200-unit purchases)
- Standardized 28 inconsistent brand names into 20 real brands
- Merged transaction and customer data (100% match rate)
- Final dataset: 249,667 rows across 71,517 customers

## 3. Key Findings

### Sales by Segment
```{r chart1, echo=FALSE}
merged_data %>%
  group_by(SEGMENT) %>%
  summarise(total_sales = sum(TOT_SALES)) %>%
  ggplot(aes(x = reorder(SEGMENT, total_sales), y = total_sales)) +
  geom_col(fill = "#D55E00") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 1/1000, suffix = "k")) +
  labs(title = "Older Families on Budget Plans Drive the Highest Chip Sales",
       x = NULL, y = "Total Sales") +
  theme_minimal(base_size = 11)
```

Older Families - Budget generates the highest total sales ($158k), 
driven by purchase frequency (4.76 transactions/customer) rather 
than basket size.

### Sales by Brand
```{r chart2, echo=FALSE}
merged_data %>%
  group_by(BRAND) %>%
  summarise(total_sales = sum(TOT_SALES)) %>%
  ggplot(aes(x = reorder(BRAND, total_sales), y = total_sales)) +
  geom_col(fill = "#D55E00") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 1/1000, suffix = "k")) +
  labs(title = "Kettle Leads Chip Sales", x = NULL, y = "Total Sales") +
  theme_minimal(base_size = 11)
```

Kettle leads all brands. Doritos sells fewer units than Smiths but 
earns more revenue, commanding a ~26% price premium per unit.

### Price Per Unit by Customer Tier
```{r chart3, echo=FALSE}
merged_data %>%
  group_by(PREMIUM_CUSTOMER) %>%
  summarise(avg_price = mean(PRICE_PER_UNIT)) %>%
  ggplot(aes(x = PREMIUM_CUSTOMER, y = avg_price)) +
  geom_col(fill = "#0072B2", width = 0.6) +
  coord_cartesian(ylim = c(0, 4.5)) +
  labs(title = "Price Per Unit Is Nearly Flat Across Tiers",
       x = NULL, y = "Avg Price Per Unit") +
  theme_minimal(base_size = 11)
```

Price per unit is nearly flat across Budget/Mainstream/Premium 
(~$3.79-$3.86). Budget customers spend more overall through higher 
purchase volume (6.96 units/customer vs Premium's 6.73), not price.

## 4. Recommendations

**1. Prioritize Older & Young Families (Budget tier).** Highest 
total sales and spend per customer, driven by frequency. Target with 
multi-buy loyalty offers matching their existing small/frequent 
purchase pattern.

**2. Don't assume Premium tier pays more for chips.** Price is flat 
across tiers; Premium marketing should focus on volume incentives, 
not price positioning.

**3. Protect Kettle and Doritos shelf space.** Doritos demonstrates 
real pricing power (~26% premium over Smiths) despite lower volume - 
worth testing modest price increases.

**4. Prioritize 175g stock reliability.** Dominant pack size overall 
and among top segments specifically - no new pack size needed, just 
availability.