---
title: "Quantium Retail Strategy & Analytics - Task 2: Trial Store Assessment"
author: "Olamilekan Olaiya"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(scales)

data <- read_csv("C:/Users/olaiy/OneDrive/Documents/Quantium/cleaned_data/QVI_features_engineered.csv")
data$DATE <- as.Date(data$DATE)
data <- data %>% mutate(YEARMONTH = format(DATE, "%Y%m"))
```

## 1. Business Problem

Julia, Category Manager for chips, ran a new store layout trial in three
stores (77, 86, 88). This analysis determines whether the trial layout
had a measurable positive effect on sales, using matched control stores
as a comparison baseline, and provides a recommendation on wider rollout.

## 2. Methodology

For each trial store, a control store was selected from the remaining
store population by combining two similarity measures calculated over
the pre-trial period (July 2018 - January 2019):

- **Pearson correlation** - how closely a candidate store's monthly
  trend shape matches the trial store's
- **Magnitude distance** - how close a candidate store's sales/customer
  scale is to the trial store's, rescaled to a 0-1 score

The two scores were averaged per metric (sales, customers), then
averaged again across both metrics to produce one overall match score
per candidate store.

## 3. Monthly Store Metrics

```{r store-metrics}
store_metrics <- data %>%
  group_by(STORE_NBR, YEARMONTH) %>%
  summarise(
    total_sales = sum(TOT_SALES),
    total_customers = n_distinct(LYLTY_CARD_NBR),
    avg_txn_per_customer = n_distinct(TXN_ID) / n_distinct(LYLTY_CARD_NBR),
    .groups = "drop"
  )
head(store_metrics)
```

## 4. Control Store Selection Functions

```{r functions}
calculate_correlation <- function(input_table, metric_col, trial_store) {
  pre_trial <- input_table %>% filter(YEARMONTH < "201902")
  trial_data <- pre_trial %>% filter(STORE_NBR == trial_store) %>%
    select(YEARMONTH, trial_metric = all_of(metric_col))
  other_stores <- pre_trial %>% filter(STORE_NBR != trial_store)
  other_stores %>%
    left_join(trial_data, by = "YEARMONTH") %>%
    group_by(STORE_NBR) %>%
    summarise(correlation = cor(.data[[metric_col]], trial_metric), .groups = "drop")
}

calculate_magnitude_distance <- function(input_table, metric_col, trial_store) {
  pre_trial <- input_table %>% filter(YEARMONTH < "201902")
  trial_data <- pre_trial %>% filter(STORE_NBR == trial_store) %>%
    select(YEARMONTH, trial_metric = all_of(metric_col))
  other_stores <- pre_trial %>% filter(STORE_NBR != trial_store)
  other_stores %>%
    left_join(trial_data, by = "YEARMONTH") %>%
    mutate(abs_diff = abs(.data[[metric_col]] - trial_metric)) %>%
    group_by(STORE_NBR) %>%
    summarise(avg_distance = mean(abs_diff), .groups = "drop") %>%
    mutate(magnitude_score = 1 - (avg_distance - min(avg_distance)) /
                                  (max(avg_distance) - min(avg_distance)))
}

select_control_store <- function(store_metrics, trial_store) {
  sales <- calculate_correlation(store_metrics, "total_sales", trial_store) %>%
    left_join(calculate_magnitude_distance(store_metrics, "total_sales", trial_store),
               by = "STORE_NBR") %>%
    mutate(sales_score = (correlation + magnitude_score) / 2) %>%
    select(STORE_NBR, sales_score)

  cust <- calculate_correlation(store_metrics, "total_customers", trial_store) %>%
    left_join(calculate_magnitude_distance(store_metrics, "total_customers", trial_store),
               by = "STORE_NBR") %>%
    mutate(cust_score = (correlation + magnitude_score) / 2) %>%
    select(STORE_NBR, cust_score)

  sales %>%
    inner_join(cust, by = "STORE_NBR") %>%
    mutate(overall_score = (sales_score + cust_score) / 2) %>%
    arrange(desc(overall_score))
}

plot_trial_vs_control <- function(store_metrics, trial_store, control_store) {
  store_metrics %>%
    filter(STORE_NBR %in% c(trial_store, control_store)) %>%
    mutate(STORE_TYPE = if_else(STORE_NBR == trial_store, "Trial", "Control")) %>%
    ggplot(aes(x = YEARMONTH, y = total_sales, color = STORE_TYPE, group = STORE_NBR)) +
    geom_line(linewidth = 1) + geom_point() +
    geom_vline(xintercept = "201902", linetype = "dashed", color = "gray50") +
    labs(title = paste("Trial Store", trial_store, "vs Control Store", control_store),
         subtitle = "Dashed line marks trial period start (Feb 2019)",
         x = "Month", y = "Total Sales", color = NULL) +
    scale_color_manual(values = c("Trial" = "#0072B2", "Control" = "#D55E00")) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 90))
}

run_trial_ttest <- function(store_metrics, trial_store, control_store) {
  trial_period <- store_metrics %>%
    filter(STORE_NBR %in% c(trial_store, control_store),
           YEARMONTH >= "201902", YEARMONTH <= "201904")
  t.test(
    trial_period$total_sales[trial_period$STORE_NBR == trial_store],
    trial_period$total_sales[trial_period$STORE_NBR == control_store],
    alternative = "greater"
  )
}
```

## 5. Trial Store 77

```{r store-77-selection}
selection_77 <- select_control_store(store_metrics, 77)
head(selection_77, 5)
```

Store 233 selected as control (overall score 0.988).

```{r store-77-chart}
plot_trial_vs_control(store_metrics, 77, 233)
```

```{r store-77-ttest}
run_trial_ttest(store_metrics, 77, 233)
```

Result: $62/month average lift, p = 0.054 - borderline, driven by
customer count growth (43 to 47) while the control store declined
(43 to 27).

## 6. Trial Store 86

```{r store-86-selection}
selection_86 <- select_control_store(store_metrics, 86)
head(selection_86, 5)
```

Store 155 selected as control (overall score 0.919).

```{r store-86-chart}
plot_trial_vs_control(store_metrics, 86, 155)
```

```{r store-86-ttest}
run_trial_ttest(store_metrics, 86, 155)
```

Result: $68/month average lift, p = 0.127 - not statistically
significant; March's spike did not sustain into April.

## 7. Trial Store 88

```{r store-88-selection}
selection_88 <- select_control_store(store_metrics, 88)
head(selection_88, 5)
```

Store 237 selected as control (overall score 0.767).

```{r store-88-chart}
plot_trial_vs_control(store_metrics, 88, 237)
```

```{r store-88-ttest}
run_trial_ttest(store_metrics, 88, 237)
```

Result: $160/month average lift, p = 0.040 - statistically
significant, driven by both more customers (122 to 133) and higher
purchase frequency per customer.

## 8. Combined Comparison

```{r all-stores-chart, fig.height=10}
comparison_data <- store_metrics %>%
  filter(STORE_NBR %in% c(77, 233, 86, 155, 88, 237)) %>%
  mutate(
    PAIR = case_when(
      STORE_NBR %in% c(77, 233) ~ "Store 77 vs Control 233",
      STORE_NBR %in% c(86, 155) ~ "Store 86 vs Control 155",
      STORE_NBR %in% c(88, 237) ~ "Store 88 vs Control 237"
    ),
    STORE_TYPE = if_else(STORE_NBR %in% c(77, 86, 88), "Trial", "Control")
  )

ggplot(comparison_data, aes(x = YEARMONTH, y = total_sales,
                             color = STORE_TYPE, group = STORE_NBR)) +
  geom_line(linewidth = 1) + geom_point() +
  geom_vline(xintercept = "201902", linetype = "dashed", color = "gray50") +
  facet_wrap(~PAIR, scales = "free_y", ncol = 1) +
  labs(title = "Trial vs Control Store Performance Across All Three Trials",
       subtitle = "Dashed line marks trial period start (Feb 2019)",
       x = "Month", y = "Total Sales", color = NULL) +
  scale_color_manual(values = c("Trial" = "#0072B2", "Control" = "#D55E00")) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
        strip.text = element_text(face = "bold"))
```

## 9. Summary of Findings

| Trial Store | Control | Avg Monthly Lift | p-value | Significant? | Primary Driver |
|---|---|---|---|---|---|
| 77 | 233 | +\$62 | 0.054 | Borderline | Customer count growth |
| 86 | 155 | +\$68 | 0.127 | No | Neither driver moved meaningfully |
| 88 | 237 | +\$160 | 0.040 | Yes | More customers and higher frequency |

## 10. Recommendations

**1. Store 88's result supports the layout change as effective and should
anchor further rollout decisions.** Statistically significant lift driven
by both customer growth and purchase frequency - the strongest, most
complete evidence in the trial.

**2. Store 77's result is promising but not yet conclusive.** The trend
is visible but narrowly misses significance, likely due to the short
3-month trial window; recommend extending the trial period to confirm.

**3. Store 86 shows no credible evidence of impact and should not be
used to justify a rollout.** Recommend investigating what differs about
this store's profile to understand why the layout didn't take effect.

**4. Do not recommend an immediate full chain-wide rollout based on this
trial alone.** Results were inconsistent across the three stores tested.
A second-phase trial with more stores and a longer test window is
recommended to identify which store characteristics predict a positive
response to the layout change.

